Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-02-05 submission/manuscript.Rmd
Insert references, closes #1

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
editor_options:
  chunk_output_type: console
---

@@

16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s. The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences. ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold. However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical. Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs. Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU. The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.

@@

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a *E. coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` *E. coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

At broad thresholds thresholds multiple species could be represented by the same OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant risk of splitting single genomes into multiple bins if too fine of a threshold is applied to defining an OTU. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference. Since there is no consensus definition for a biological species concept, microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner. This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names. Furthermore, it is unlikely that the 16S rRNA gene evolves at the same rate across all bacterial lineages, which limits the biological interpretation of a common OTU definition. At best, a distance-based definition of a taxonomic unit is operational. There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data (CITATION). We are asking too much of a short section of a bacterial genome to be able to differentiate between species. It is difficult to defend a unit of inference that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possability that parts of a genome would have different ecologies. Although there are multiple reasons that proponents of ASVs encourage their use, the significant risk of splitting genomes is too high to warrant their use.

**Materials and Methods. (i) Data availability.** The 16S rRNA gene sequences used in this study were  obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019). At the time of submission, this is the most current version of the database. The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database. The *rrn*DB provides downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy. For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical. Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome. Because strain names were not consistently given to genomes across bacterial species, the strain level designations were ignored.

**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v138) using the mothur software package (v. 1.XX). Regions of the 16S rRNA gene were selected because of their use in the microbial ecology literature. Full length sequences corresponded to *E. coli* positions XX through XXXX, V4 to positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5 to positions XXX through XXX.

@@

>>>>

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
editor_options:
  chunk_output_type: console
bibliography: references.bib
csl: asm.csl
---

@@

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s. The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017]. However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical [@Pei2010; @Sun2013]. Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs. Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU. The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.

@@

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB)[@Stoddard2014]. Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g., *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a *E. coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` *E. coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

At broad thresholds multiple species could be represented by the same OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant risk of splitting single genomes into multiple bins if too fine of a threshold is applied to defining an OTU. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference [@Callahan2017; @Schloss2011; @Johnson2019]. Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016]). This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005]. Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition. A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014]. There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data [@Staley2006; @Oren2013; @Baltrus2016]. We are asking too much of a short section of a bacterial genome to be able to differentiate between species. It is difficult to defend a unit of inference that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possibility that parts of a genome would have different ecologies. Although there are multiple reasons that proponents of ASVs encourage their use, the significant risk of splitting genomes is too high to warrant their use.

**Materials and Methods. (i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019) [@Stoddard2014]. At the time of submission, this is the most current version of the database. The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database. The *rrn*DB provides downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy. For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical. Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome. Because strain names were not consistently given to genomes across bacterial species, the strain level designations were ignored.

**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v138) using the mothur software package (v. 1.XX) [@Schloss2009; @Quast2012]. Regions of the 16S rRNA gene were selected because of their use in the microbial ecology literature. Full length sequences corresponded to *E. coli* positions XX through XXXX, V4 to positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5 to positions XXX through XXX.

@@

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
